---
title: "Stem-like CD8 T-cell Analysis — Initial Report"
subtitle: "Cell recovery and inclusive co-expression of the requested stem-like CD8 markers"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-tools: true
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
params:
  rds_path: "obj_mod.rds"
---

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(Seurat)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(fs)
  library(ComplexHeatmap)
  library(circlize)
  library(grid)
  library(knitr)
})

# ============================================================
# INPUT — taken directly from the existing analysis script
# ============================================================

rds_path <- params$rds_path

message("Reading: ", rds_path)

if (!file.exists(rds_path)) {
  stop(
    paste0(
      "Cannot find Seurat object: ", rds_path,
      "\nThe original analysis script uses obj_mod.rds by default.",
      "\nChange params:rds_path in this QMD if the object is elsewhere."
    )
  )
}

obj <- readRDS(rds_path)

OUTDIR <- "CD8_STEM_INITIAL_REPORT"
dir_create(OUTDIR)

# ============================================================
# EXPERIMENTAL DESIGN — copied from the existing script
# ============================================================

untreated_lanes <- c(
  1, 3, 5, 7,
  33, 35, 37, 39
)

treated_lanes <- c(
  2, 4, 6, 8,
  34, 36, 38, 40
)

untrained_lanes <- c(1, 2, 33, 34)
SYKi_lanes      <- c(3, 4, 35, 36)
MDP_lanes       <- c(5, 6, 37, 38)
BG_lanes        <- c(7, 8, 39, 40)

# ============================================================
# REQUIRED METADATA
# ============================================================

if (!"Lane" %in% colnames(obj@meta.data)) {
  stop("The object does not contain a metadata column named 'Lane'.")
}

if (!"predicted.celltype.l1" %in% colnames(obj@meta.data)) {
  stop("The object does not contain predicted.celltype.l1.")
}

# ============================================================
# NORMALIZE LANE NUMBER
# ============================================================

lane_character <- as.character(obj$Lane)

lane_numeric <- suppressWarnings(
  as.numeric(lane_character)
)

if (all(is.na(lane_numeric))) {
  lane_numeric <- suppressWarnings(
    as.numeric(
      gsub("\\D+", "", lane_character)
    )
  )
}

obj$Lane_numeric <- lane_numeric

# ============================================================
# CREATE TREATMENT / TRAINING METADATA
# ============================================================

obj$Treatment <- case_when(
  obj$Lane_numeric %in% untreated_lanes ~ "Untreated",
  obj$Lane_numeric %in% treated_lanes   ~ "Treated",
  TRUE ~ NA_character_
)

obj$LaneGroup <- case_when(
  obj$Lane_numeric %in% untrained_lanes ~ "untrained",
  obj$Lane_numeric %in% SYKi_lanes      ~ "SYKi",
  obj$Lane_numeric %in% MDP_lanes       ~ "MDP",
  obj$Lane_numeric %in% BG_lanes        ~ "BG",
  TRUE ~ NA_character_
)

obj$Treatment <- factor(
  obj$Treatment,
  levels = c("Untreated", "Treated")
)

obj$LaneGroup <- factor(
  obj$LaneGroup,
  levels = c("untrained", "SYKi", "MDP", "BG")
)

obj$LaneGroup_Treatment <- paste(
  as.character(obj$LaneGroup),
  as.character(obj$Treatment),
  sep = "_"
)

obj$LaneGroup_Treatment <- factor(
  obj$LaneGroup_Treatment,
  levels = c(
    "untrained_Untreated",
    "untrained_Treated",
    "SYKi_Untreated",
    "SYKi_Treated",
    "MDP_Untreated",
    "MDP_Treated",
    "BG_Untreated",
    "BG_Treated"
  )
)

# ============================================================
# GENES
# ============================================================

stem_genes <- c(
  "TCF7",
  "PDCD1",
  "SLAMF6",
  "CD28",
  "CCR7",
  "SELL"
)

comparison_gene <- "IFNG"
all_genes <- c(stem_genes, comparison_gene)

# ============================================================
# IDENTIFY / SUBSET CD8 CELLS
# ============================================================

cd8_labels <- unique(
  as.character(
    obj$predicted.celltype.l1[
      grepl(
        "CD8",
        obj$predicted.celltype.l1,
        ignore.case = TRUE
      )
    ]
  )
)

cd8_labels <- cd8_labels[!is.na(cd8_labels)]

if (length(cd8_labels) == 0) {
  stop("No CD8 labels were found in predicted.celltype.l1.")
}

cd8 <- subset(
  obj,
  subset = predicted.celltype.l1 %in% cd8_labels
)

if (ncol(cd8) == 0) {
  stop("CD8 subset contains zero cells.")
}

DefaultAssay(cd8) <- "RNA"

# ============================================================
# CHECK REQUESTED GENES
# ============================================================

gene_check <- data.frame(
  Gene = all_genes,
  Present = all_genes %in% rownames(cd8[["RNA"]])
)

missing_genes <- gene_check$Gene[!gene_check$Present]

if (length(missing_genes) > 0) {
  stop(
    "Missing genes: ",
    paste(missing_genes, collapse = ", ")
  )
}

# ============================================================
# CREATE CELL-LEVEL DATA — same approach as original script
# ============================================================

cell_data <- FetchData(
  cd8,
  vars = c(
    all_genes,
    "Lane_numeric",
    "LaneGroup",
    "Treatment",
    "LaneGroup_Treatment"
  )
)

GROUP_COL <- "LaneGroup_Treatment"

TREATMENT_LEVELS <- c(
  "untrained_Untreated",
  "untrained_Treated",
  "SYKi_Untreated",
  "SYKi_Treated",
  "MDP_Untreated",
  "MDP_Treated",
  "BG_Untreated",
  "BG_Treated"
)

cell_data[[GROUP_COL]] <- factor(
  cell_data[[GROUP_COL]],
  levels = TREATMENT_LEVELS
)

combo_binary <- cell_data %>%
  filter(
    !is.na(.data[[GROUP_COL]]),
    .data[[GROUP_COL]] %in% TREATMENT_LEVELS
  ) %>%
  mutate(
    across(all_of(stem_genes), ~ .x > 0)
  )
```

# Overview

The question is whether the CD8 scRNA-seq data contain cells with a transcriptional phenotype corresponding to the stem-like markers requested in the email:

**TCF1+, PD-1+, SLAMF6+, CD28+**, with possible **CCR7+ and/or CD62L+** characteristics.

For the scRNA-seq analysis, the corresponding genes are:

| Protein marker | RNA proxy |
|---|---|
| TCF1 | **TCF7** |
| PD-1 | **PDCD1** |
| SLAMF6 | **SLAMF6** |
| CD28 | **CD28** |
| CCR7 | **CCR7** |
| CD62L | **SELL** |

The purpose of this initial report is to determine:

1. whether sufficient CD8 cells were recovered in each experimental condition;
2. how many cells satisfy the requested marker combinations using **inclusive co-expression**;
3. whether the requested multi-marker populations are sufficiently represented for group-level analysis.

> **Practical analysis criterion:** For this type of cell-population comparison, we typically prefer **more than 100 cells per cell population in each group under analysis**. Populations substantially below this level are considered too sparsely represented for a robust group-level comparison in this initial analysis.

A further biological consideration is that several of these markers reflect a naïve/stem-like state; **PMA strongly activates T cells and can downregulate or alter this transcriptional program**, so loss of the RNA-defined combination after stimulation does not necessarily mean loss of the underlying cells.

> **Important limitation:** CITE-seq was not performed for these markers. The analysis therefore measures RNA co-detection rather than the protein-defined TCF1+ PD-1+ SLAMF6+ CD28+ phenotype.

# 1. Cell recovery before phenotype interpretation

The original analysis correctly examines cell recovery before interpreting phenotype frequencies. Lower numbers of recovered cells should not automatically be interpreted as cell death.

```{r}
#| label: recovery-analysis

all_cell_recovery_lane <- obj@meta.data %>%
  filter(
    !is.na(Lane_numeric),
    !is.na(Treatment),
    !is.na(LaneGroup)
  ) %>%
  group_by(
    Lane_numeric,
    LaneGroup,
    Treatment
  ) %>%
  summarise(
    Total_cells = n(),
    .groups = "drop"
  )

cd8_cell_recovery_lane <- cd8@meta.data %>%
  filter(
    !is.na(Lane_numeric),
    !is.na(Treatment),
    !is.na(LaneGroup)
  ) %>%
  group_by(
    Lane_numeric,
    LaneGroup,
    Treatment
  ) %>%
  summarise(
    CD8_cells = n(),
    .groups = "drop"
  )

recovery_by_lane <- all_cell_recovery_lane %>%
  left_join(
    cd8_cell_recovery_lane,
    by = c(
      "Lane_numeric",
      "LaneGroup",
      "Treatment"
    )
  ) %>%
  mutate(
    CD8_cells = ifelse(is.na(CD8_cells), 0, CD8_cells),
    Percent_CD8 = 100 * CD8_cells / Total_cells
  )

all_recovery_summary <- obj@meta.data %>%
  filter(
    !is.na(Treatment),
    !is.na(LaneGroup)
  ) %>%
  group_by(
    LaneGroup,
    Treatment
  ) %>%
  summarise(
    Total_cells = n(),
    .groups = "drop"
  )

cd8_recovery_summary <- cd8@meta.data %>%
  filter(
    !is.na(Treatment),
    !is.na(LaneGroup)
  ) %>%
  group_by(
    LaneGroup,
    Treatment
  ) %>%
  summarise(
    CD8_cells = n(),
    .groups = "drop"
  )

recovery_summary <- all_recovery_summary %>%
  left_join(
    cd8_recovery_summary,
    by = c("LaneGroup", "Treatment")
  ) %>%
  mutate(
    CD8_cells = ifelse(is.na(CD8_cells), 0, CD8_cells),
    Percent_CD8 = 100 * CD8_cells / Total_cells
  )

write.csv(
  recovery_by_lane,
  file.path(OUTDIR, "00_cell_recovery_by_lane.csv"),
  row.names = FALSE
)

write.csv(
  recovery_summary,
  file.path(OUTDIR, "00_cell_recovery_summary.csv"),
  row.names = FALSE
)

kable(
  recovery_summary,
  digits = 1,
  caption = "Recovered cells and CD8 representation by condition"
)
```

```{r}
#| label: cd8-recovery-plot
#| fig-width: 11
#| fig-height: 6

p_recovery_summary <- ggplot(
  recovery_summary,
  aes(
    x = LaneGroup,
    y = CD8_cells,
    fill = Treatment
  )
) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +
  geom_text(
    aes(
      label = paste0(
        format(CD8_cells, big.mark = ","),
        "\n(",
        round(Percent_CD8, 1),
        "% of recovered)"
      )
    ),
    position = position_dodge(width = 0.8),
    vjust = -0.25,
    size = 3.5
  ) +
  labs(
    title = "CD8 cell recovery before phenotype analysis",
    subtitle = "Absolute CD8 N with percentage of all recovered cells",
    x = "Training condition",
    y = "Number of recovered CD8 cells",
    fill = "Treatment"
  ) +
  theme_classic(base_size = 14) +
  expand_limits(
    y = max(recovery_summary$CD8_cells, na.rm = TRUE) * 1.18
  )

p_recovery_summary
```

### Comment on cell recovery

The treated samples contain consistently fewer recovered CD8 cells than the corresponding untreated samples. In the current data, untreated groups contain approximately **8,300–9,100 CD8 cells**, whereas treated groups contain approximately **2,800–3,600 CD8 cells**. CD8 cells also represent a lower fraction of all recovered cells in treated samples (**13.4–15.3%**) than in untreated samples (**32.6–36.6%**).

This difference is important for the interpretation of the co-expression analysis. **Absolute numbers of marker-positive cells should not be compared by themselves across untreated and treated groups.** The percentage of CD8 cells carrying each phenotype is therefore shown together with the cell count in the heatmap. At this stage, the recovery difference should be treated as a sampling/recovery consideration rather than interpreted as evidence of CD8-cell loss.

# 2. Inclusive co-expression heatmap

This is the main first-report figure.

These are **inclusive combinations**. For example:

**TCF7 + CD28 + CCR7 + SELL**

means that a cell is positive for all four transcripts, regardless of whether **PDCD1** or **SLAMF6** are also detected.

Therefore:

**n(TCF7 + CD28 + CCR7 + SELL) ≤ n(TCF7)**

must always be true.

```{r}
#| label: inclusive-heatmap
#| fig-width: 13
#| fig-height: 10

inclusive_combos <- list(
  "TCF7" =
    c("TCF7"),

  "PDCD1" =
    c("PDCD1"),

  "SLAMF6" =
    c("SLAMF6"),

  "CD28" =
    c("CD28"),

  "CCR7" =
    c("CCR7"),

  "SELL" =
    c("SELL"),

  "TCF7 + SLAMF6" =
    c("TCF7", "SLAMF6"),

  "TCF7 + CD28" =
    c("TCF7", "CD28"),

  "TCF7 + CCR7" =
    c("TCF7", "CCR7"),

  "TCF7 + SELL" =
    c("TCF7", "SELL"),

  "TCF7 + PDCD1" =
    c("TCF7", "PDCD1"),

  "TCF7 + SLAMF6 + CD28" =
    c("TCF7", "SLAMF6", "CD28"),

  "TCF7 + CD28 + CCR7" =
    c("TCF7", "CD28", "CCR7"),

  "TCF7 + CD28 + SELL" =
    c("TCF7", "CD28", "SELL"),

  "TCF7 + CCR7 + SELL" =
    c("TCF7", "CCR7", "SELL"),

  "TCF7 + SLAMF6 + CCR7" =
    c("TCF7", "SLAMF6", "CCR7"),

  "TCF7 + SLAMF6 + SELL" =
    c("TCF7", "SLAMF6", "SELL"),

  "TCF7 + SLAMF6 + CD28 + CCR7" =
    c("TCF7", "SLAMF6", "CD28", "CCR7"),

  "TCF7 + SLAMF6 + CD28 + SELL" =
    c("TCF7", "SLAMF6", "CD28", "SELL"),

  "TCF7 + SLAMF6 + CD28 + CCR7 + SELL" =
    c("TCF7", "SLAMF6", "CD28", "CCR7", "SELL"),

  "TCF7 + PDCD1 + SLAMF6 + CD28" =
    c("TCF7", "PDCD1", "SLAMF6", "CD28"),

  "TCF7 + PDCD1 + SLAMF6 + CD28 + CCR7" =
    c("TCF7", "PDCD1", "SLAMF6", "CD28", "CCR7"),

  "TCF7 + PDCD1 + SLAMF6 + CD28 + SELL" =
    c("TCF7", "PDCD1", "SLAMF6", "CD28", "SELL"),

  "TCF7 + PDCD1 + SLAMF6 + CD28 + CCR7 + SELL" =
    c("TCF7", "PDCD1", "SLAMF6", "CD28", "CCR7", "SELL")
)

count_inclusive_combo <- function(df, markers) {
  marker_matrix <- as.matrix(
    df[, markers, drop = FALSE]
  )

  sum(
    rowSums(marker_matrix) ==
      length(markers)
  )
}

inclusive_list <- lapply(
  TREATMENT_LEVELS,
  function(grp) {

    df_grp <- combo_binary %>%
      filter(
        .data[[GROUP_COL]] == grp
      )

    total_cells <- nrow(df_grp)

    counts <- vapply(
      inclusive_combos,
      function(markers) {
        count_inclusive_combo(
          df_grp,
          markers
        )
      },
      numeric(1)
    )

    data.frame(
      Treatment = grp,
      Combination = names(counts),
      Cells = as.numeric(counts),
      Percent_CD8 =
        if (total_cells > 0) {
          100 * as.numeric(counts) / total_cells
        } else {
          NA_real_
        },
      Total_CD8 = total_cells,
      stringsAsFactors = FALSE
    )
  }
)

inclusive_long <- bind_rows(
  inclusive_list
)

# ============================================================
# SANITY CHECK
# Every TCF7-containing combination must be <= total TCF7+
# ============================================================

sanity_check <- inclusive_long %>%
  group_by(Treatment) %>%
  mutate(
    TCF7_total =
      Cells[
        Combination == "TCF7"
      ]
  ) %>%
  filter(
    grepl(
      "TCF7",
      Combination
    )
  ) %>%
  mutate(
    Valid =
      Cells <= TCF7_total
  ) %>%
  ungroup()

if (any(!sanity_check$Valid)) {
  stop(
    paste(
      "ERROR:",
      "An inclusive combination containing TCF7",
      "has more cells than total TCF7+ cells."
    )
  )
}

write.csv(
  inclusive_long,
  file.path(
    OUTDIR,
    "03_inclusive_marker_combinations_all_groups.csv"
  ),
  row.names = FALSE
)

combo_totals <- inclusive_long %>%
  group_by(Combination) %>%
  summarise(
    Total_Cells_AllGroups =
      sum(Cells),
    .groups = "drop"
  ) %>%
  arrange(
    desc(Total_Cells_AllGroups)
  )

HEATMAP_TOP_N_COMBOS <- 20

top_combos <- head(
  combo_totals$Combination,
  HEATMAP_TOP_N_COMBOS
)

mat2_pct_df <- inclusive_long %>%
  filter(
    Combination %in% top_combos
  ) %>%
  select(
    Treatment,
    Combination,
    Percent_CD8
  ) %>%
  pivot_wider(
    names_from = Treatment,
    values_from = Percent_CD8,
    values_fill = 0
  )

mat2_cells_df <- inclusive_long %>%
  filter(
    Combination %in% top_combos
  ) %>%
  select(
    Treatment,
    Combination,
    Cells
  ) %>%
  pivot_wider(
    names_from = Treatment,
    values_from = Cells,
    values_fill = 0
  )

mat2_pct_df <- mat2_pct_df[
  match(
    top_combos,
    mat2_pct_df$Combination
  ),
]

mat2_cells_df <- mat2_cells_df[
  match(
    top_combos,
    mat2_cells_df$Combination
  ),
]

group_cols <- intersect(
  TREATMENT_LEVELS,
  colnames(mat2_pct_df)
)

mat2 <- as.matrix(
  mat2_pct_df[
    ,
    group_cols,
    drop = FALSE
  ]
)

mat2_cells <- as.matrix(
  mat2_cells_df[
    ,
    group_cols,
    drop = FALSE
  ]
)

rownames(mat2) <-
  mat2_pct_df$Combination

rownames(mat2_cells) <-
  mat2_cells_df$Combination

max_pct <- max(
  mat2,
  na.rm = TRUE
)

if (
  !is.finite(max_pct) ||
  max_pct <= 0
) {
  max_pct <- 1
}

col_fun2 <- circlize::colorRamp2(
  c(
    0,
    max_pct / 2,
    max_pct
  ),
  c(
    "white",
    "steelblue",
    "darkblue"
  )
)

ht2 <- Heatmap(
  mat2,
  name = "% of group",
  col = col_fun2,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_names_side = "left",
  row_names_max_width =
    unit(11, "cm"),
  column_title =
    paste0(
      "Top ",
      nrow(mat2),
      " INCLUSIVE marker combinations (% and n of group's CD8 cells)"
    ),
  column_title_gp =
    gpar(
      fontsize = 14,
      fontface = "bold"
    ),
  cell_fun =
    function(
      j,
      i,
      x,
      y,
      width,
      height,
      fill
    ) {

      grid.text(
        paste0(
          format(
            mat2_cells[i, j],
            big.mark = ","
          ),
          "\n(",
          sprintf(
            "%.1f",
            mat2[i, j]
          ),
          "%)"
        ),
        x,
        y,
        gp =
          gpar(
            fontsize = 6.5,
            lineheight = 0.85
          )
      )
    },
  rect_gp =
    gpar(
      col = "grey80",
      lwd = 0.5
    )
)

draw(ht2)

write.csv(
  mat2_cells_df %>%
    rename_with(
      ~ paste0("Cells_", .x),
      all_of(group_cols)
    ) %>%
    left_join(
      mat2_pct_df %>%
        rename_with(
          ~ paste0("Pct_", .x),
          all_of(group_cols)
        ),
      by = "Combination"
    ),
  file.path(
    OUTDIR,
    "03_inclusive_heatmap_data.csv"
  ),
  row.names = FALSE
)
```

### Comment on inclusive co-expression

The heatmap separates two important observations.

First, the **TCF7 + SLAMF6 + CD28 + CCR7 + SELL** phenotype **without PDCD1** is well represented in every untreated group: **482 cells (5.3%) in untrained, 629 (7.2%) in SYKi, 494 (5.9%) in MDP, and 382 (4.6%) in BG**. These untreated populations are above the **>100 cells per population per group** that we typically prefer for this type of group-level analysis. SYKi_Untreated has the highest observed representation of this five-marker phenotype, although this heatmap alone does not establish that the difference is statistically or biologically significant.

Second, this same five-marker phenotype is essentially absent after treatment/PMA (**0, 0, 1, and 0 cells**, respectively). In contrast, **PDCD1 is more frequently detected in the treated groups**, while CCR7/SELL-containing stem-like or naïve-associated combinations collapse after stimulation. Thus, the limiting feature for the **originally requested phenotype** is the requirement for simultaneous **PDCD1** expression with the TCF7/SLAMF6/CD28/CCR7/SELL program, not a general absence of the stem-like/naïve-associated cells before stimulation.

**PMA is a strong activation stimulus and is expected to alter/downregulate naïve or stem-like transcriptional features such as CCR7/SELL/TCF7, so the near-loss of this RNA-defined phenotype after stimulation should not be interpreted as physical disappearance of the underlying cells.**

# 3. Summary

This section summarizes the main first-pass findings from the cell-recovery analysis and the inclusive co-expression analysis above.

```{r}
#| label: v1-summary
#| results: asis

five_no_pdcd1_name <-
  "TCF7 + SLAMF6 + CD28 + CCR7 + SELL"

core_name <-
  "TCF7 + PDCD1 + SLAMF6 + CD28"

core_ccr7_name <-
  "TCF7 + PDCD1 + SLAMF6 + CD28 + CCR7"

core_sell_name <-
  "TCF7 + PDCD1 + SLAMF6 + CD28 + SELL"

all6_name <-
  "TCF7 + PDCD1 + SLAMF6 + CD28 + CCR7 + SELL"

five_no_pdcd1_tbl <- inclusive_long %>%
  filter(Combination == five_no_pdcd1_name)

core_tbl <- inclusive_long %>%
  filter(Combination == core_name)

core_ccr7_tbl <- inclusive_long %>%
  filter(Combination == core_ccr7_name)

core_sell_tbl <- inclusive_long %>%
  filter(Combination == core_sell_name)

all6_tbl <- inclusive_long %>%
  filter(Combination == all6_name)

cat(
  "A total of **",
  format(ncol(cd8), big.mark = ","),
  " CD8 cells** were identified from `predicted.celltype.l1` and included in this initial analysis.\n\n",
  sep = ""
)

if (nrow(five_no_pdcd1_tbl) > 0) {
  untreated_five <- five_no_pdcd1_tbl %>%
    filter(grepl("_Untreated$", Treatment))

  treated_five <- five_no_pdcd1_tbl %>%
    filter(grepl("_Treated$", Treatment))

  cat(
    "The **TCF7 + SLAMF6 + CD28 + CCR7 + SELL** phenotype without PDCD1 is well represented before PMA/treatment, ",
    "with **",
    min(untreated_five$Cells, na.rm = TRUE),
    " to ",
    max(untreated_five$Cells, na.rm = TRUE),
    " cells per untreated group**. ",
    "After treatment/PMA, the same phenotype falls to **",
    min(treated_five$Cells, na.rm = TRUE),
    " to ",
    max(treated_five$Cells, na.rm = TRUE),
    " cells per group**.\n\n",
    sep = ""
  )
}

if (nrow(core_tbl) > 0) {
  cat(
    "For the requested RNA-defined core phenotype (**TCF7 + PDCD1 + SLAMF6 + CD28**), ",
    "group-level frequencies range from **",
    sprintf("%.3f", min(core_tbl$Percent_CD8, na.rm = TRUE)),
    "% to ",
    sprintf("%.3f", max(core_tbl$Percent_CD8, na.rm = TRUE)),
    "%**, corresponding to **",
    min(core_tbl$Cells, na.rm = TRUE),
    " to ",
    max(core_tbl$Cells, na.rm = TRUE),
    " cells per group**.\n\n",
    sep = ""
  )
}

if (nrow(core_ccr7_tbl) > 0) {
  cat(
    "When **CCR7** is added to the requested four-marker phenotype, the observed frequencies range from **",
    sprintf("%.3f", min(core_ccr7_tbl$Percent_CD8, na.rm = TRUE)),
    "% to ",
    sprintf("%.3f", max(core_ccr7_tbl$Percent_CD8, na.rm = TRUE)),
    "%** across groups.\n\n",
    sep = ""
  )
}

if (nrow(core_sell_tbl) > 0) {
  cat(
    "When **SELL (CD62L)** is added to the requested four-marker phenotype, the observed frequencies range from **",
    sprintf("%.3f", min(core_sell_tbl$Percent_CD8, na.rm = TRUE)),
    "% to ",
    sprintf("%.3f", max(core_sell_tbl$Percent_CD8, na.rm = TRUE)),
    "%** across groups.\n\n",
    sep = ""
  )
}

if (nrow(all6_tbl) > 0) {
  cat(
    "When all six transcripts are required simultaneously, group-level frequencies range from **",
    sprintf("%.3f", min(all6_tbl$Percent_CD8, na.rm = TRUE)),
    "% to ",
    sprintf("%.3f", max(all6_tbl$Percent_CD8, na.rm = TRUE)),
    "%**, corresponding to **",
    min(all6_tbl$Cells, na.rm = TRUE),
    " to ",
    max(all6_tbl$Cells, na.rm = TRUE),
    " cells per group**.\n\n",
    sep = ""
  )
}

cat(
  "For this type of group-level cell-population analysis, we typically prefer **more than 100 cells per population in each group under analysis**. ",
  "The five-marker **TCF7 + SLAMF6 + CD28 + CCR7 + SELL** population meets this practical criterion in every untreated group, ",
  "but it is essentially absent after treatment/PMA. ",
  "The originally requested phenotype that also requires **PDCD1** is much more sparsely represented and does not provide sufficient cell numbers for a robust multi-group comparison. ",
  "Therefore, the data support a sizeable pre-stimulation TCF7/SLAMF6/CD28/CCR7/SELL population, while simultaneous PDCD1 expression is the limiting feature of the requested phenotype. ",
  "Because PMA strongly alters activation state and naïve/stem-like transcriptional markers, the treated-versus-untreated difference should be interpreted as a change in RNA-defined phenotype rather than evidence that the cells themselves disappeared."
)
```

# Interpretation limitation

Because these are RNA measurements after stimulation and not CITE-seq protein measurements, the results should initially be described as a **stem-like CD8-associated transcriptional phenotype**. A low count for a strict multi-marker combination may reflect true rarity, sparse transcript detection, stimulation-associated transcriptional changes, or a combination of these factors.
