---
title: "CD8 Stem-like Analysis — Step 2: Subject-level Statistics and Marker Expression"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-tools: true
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(Seurat)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(edgeR)
  library(limma)
  library(knitr)
})

args <- commandArgs(trailingOnly = TRUE)
rds_path <- if (length(args) >= 1) args[1] else "obj_mod.rds"

message("Reading: ", rds_path)
obj <- readRDS(rds_path)

# Check required metadata
if (!"Lane" %in% colnames(obj@meta.data)) {
  stop("The object does not contain Lane.")
}

if (!"predicted.celltype.l1" %in% colnames(obj@meta.data)) {
  stop("The object does not contain predicted.celltype.l1.")
}

if (!"sCALL" %in% colnames(obj@meta.data)) {
  stop("The object does not contain subject column sCALL.")
}

# Lane definitions
untreated_lanes <- c(1, 3, 5, 7, 33, 35, 37, 39)
treated_lanes   <- c(2, 4, 6, 8, 34, 36, 38, 40)

untrained_lanes <- c(1, 2, 33, 34)
SYKi_lanes      <- c(3, 4, 35, 36)
MDP_lanes       <- c(5, 6, 37, 38)
BG_lanes        <- c(7, 8, 39, 40)

# Convert Lane to numeric
obj$Lane_numeric <- suppressWarnings(
  as.numeric(as.character(obj$Lane))
)

# Treatment assignment
obj$Treatment <- case_when(
  obj$Lane_numeric %in% untreated_lanes ~ "Untreated",
  obj$Lane_numeric %in% treated_lanes   ~ "Treated",
  TRUE ~ NA_character_
)

# Training assignment
obj$LaneGroup <- case_when(
  obj$Lane_numeric %in% untrained_lanes ~ "untrained",
  obj$Lane_numeric %in% SYKi_lanes      ~ "SYKi",
  obj$Lane_numeric %in% MDP_lanes       ~ "MDP",
  obj$Lane_numeric %in% BG_lanes        ~ "BG",
  TRUE ~ NA_character_
)

obj$Treatment <- factor(
  obj$Treatment,
  levels = c("Untreated", "Treated")
)

obj$LaneGroup <- factor(
  obj$LaneGroup,
  levels = c("untrained", "SYKi", "MDP", "BG")
)

# Identify Monaco-defined CD8 labels
cd8_labels <- unique(
  obj$predicted.celltype.l1[
    grepl(
      "CD8",
      obj$predicted.celltype.l1,
      ignore.case = TRUE
    )
  ]
)

cat(
  "CD8 labels used:",
  paste(cd8_labels, collapse = ", "),
  "\n"
)

# Subset CD8 cells
cd8 <- subset(
  obj,
  subset = predicted.celltype.l1 %in% cd8_labels
)

DefaultAssay(cd8) <- "RNA"

cat(
  "Total Monaco-defined CD8 cells:",
  format(ncol(cd8), big.mark = ","),
  "\n"
)

# Predefined marker panel
genes_to_test <- c(
  "TCF7",
  "PDCD1",
  "SLAMF6",
  "CD28",
  "CCR7",
  "SELL"
)

# Five-transcript population excluding PDCD1
five_genes <- c(
  "TCF7",
  "SLAMF6",
  "CD28",
  "CCR7",
  "SELL"
)

trained_conditions <- c(
  "SYKi",
  "MDP",
  "BG"
)

# Check genes
missing_genes <- setdiff(
  genes_to_test,
  rownames(cd8[["RNA"]])
)

if (length(missing_genes) > 0) {
  stop(
    "Missing requested RNA genes: ",
    paste(missing_genes, collapse = ", ")
  )
}
```

# 1. Statistical analysis of the five-transcript CD8 population

This analysis tests whether the proportion of Monaco-defined CD8 cells with simultaneous detection of **TCF7, SLAMF6, CD28, CCR7, and SELL RNA** differs between untrained cells and each of the three pre-PMA training conditions (SYKi, MDP, BG).

```{r}
#| label: cell-proportion-data
#| fig-width: 8
#| fig-height: 5

expr <- FetchData(
  cd8,
  vars = c(
    five_genes,
    "sCALL",
    "Treatment",
    "LaneGroup"
  )
)

prepma_subject <- expr %>%
  filter(
    Treatment == "Untreated",
    LaneGroup %in% c("untrained", trained_conditions),
    !is.na(sCALL)
  ) %>%
  mutate(
    Five_marker =
      rowSums(
        as.matrix(
          across(all_of(five_genes)) > 0
        )
      ) == length(five_genes)
  ) %>%
  group_by(
    sCALL,
    LaneGroup
  ) %>%
  summarise(
    Total_CD8 = n(),
    Five_marker_cells = sum(Five_marker),
    Percent_CD8 = 100 * Five_marker_cells / Total_CD8,
    .groups = "drop"
  ) %>%
  mutate(
    LaneGroup = factor(
      LaneGroup,
      levels = c(
        "untrained",
        trained_conditions
      )
    )
  )

kable(
  prepma_subject,
  digits = 2,
  caption = "Pre-PMA five-transcript CD8 population by subject"
)

ggplot(
  prepma_subject,
  aes(
    x = LaneGroup,
    y = Percent_CD8,
    group = sCALL
  )
) +
  geom_line(alpha = 0.65) +
  geom_point(size = 3) +
  labs(
    title = "Pre-PMA five-transcript CD8 population by subject",
    subtitle = "Simultaneous TCF7, SLAMF6, CD28, CCR7, and SELL RNA detection",
    x = NULL,
    y = "% of Monaco-defined CD8 cells"
  ) +
  theme_bw(base_size = 12)
```

# 2. Overall pre-PMA comparison

```{r}
#| label: cell-proportion-overall-friedman

complete_subjects <- prepma_subject %>%
  select(
    sCALL,
    LaneGroup,
    Percent_CD8
  ) %>%
  pivot_wider(
    names_from = LaneGroup,
    values_from = Percent_CD8
  ) %>%
  filter(
    complete.cases(
      untrained,
      SYKi,
      MDP,
      BG
    )
  )

cat(
  "Subjects with complete data for all four conditions:",
  nrow(complete_subjects),
  "\n"
)

if (nrow(complete_subjects) >= 2) {
  
  friedman_long <- complete_subjects %>%
    pivot_longer(
      cols = c(
        untrained,
        SYKi,
        MDP,
        BG
      ),
      names_to = "Condition",
      values_to = "Percent_CD8"
    )
  
  friedman_result <- friedman.test(
    Percent_CD8 ~ Condition | sCALL,
    data = friedman_long
  )
  
  cat(
    "Overall Friedman test across the four pre-PMA conditions: p = ",
    signif(friedman_result$p.value, 4),
    "\n",
    sep = ""
  )
  
} else {
  
  cat(
    "Not enough complete subjects for the Friedman test.\n"
  )
}
```

## Paired comparisons versus untrained

```{r}
#| label: cell-proportion-paired-comparisons

run_proportion_paired_test <- function(
  condition,
  baseline = "untrained"
) {
  
  pair_df <- prepma_subject %>%
    filter(
      LaneGroup %in% c(
        baseline,
        condition
      )
    ) %>%
    select(
      sCALL,
      LaneGroup,
      Percent_CD8
    ) %>%
    pivot_wider(
      names_from = LaneGroup,
      values_from = Percent_CD8
    ) %>%
    filter(
      complete.cases(
        .data[[baseline]],
        .data[[condition]]
      )
    ) %>%
    mutate(
      Difference_pct_points =
        .data[[condition]] -
        .data[[baseline]]
    )
  
  if (nrow(pair_df) < 2) {
    
    return(
      list(
        condition = condition,
        pair_df = pair_df,
        wilcox = NULL,
        limma = NULL
      )
    )
  }
  
  wilcox_result <- wilcox.test(
    pair_df[[condition]],
    pair_df[[baseline]],
    paired = TRUE,
    exact = FALSE
  )
  
  limma_df <- prepma_subject %>%
    filter(
      LaneGroup %in% c(
        baseline,
        condition
      )
    ) %>%
    mutate(
      sCALL = factor(sCALL),
      LaneGroup = factor(
        LaneGroup,
        levels = c(
          baseline,
          condition
        )
      )
    ) %>%
    arrange(
      sCALL,
      LaneGroup
    )
  
  limma_design <- model.matrix(
    ~ sCALL + LaneGroup,
    data = limma_df
  )
  
  limma_y <- matrix(
    limma_df$Percent_CD8,
    nrow = 1,
    dimnames = list(
      "Five_transcript_percent",
      NULL
    )
  )
  
  limma_fit <- limma::lmFit(
    limma_y,
    limma_design
  )
  
  limma_fit <- limma::eBayes(
    limma_fit
  )
  
  coef_name <- paste0(
    "LaneGroup",
    condition
  )
  
  limma_result <- limma::topTable(
    limma_fit,
    coef = coef_name,
    number = Inf,
    sort.by = "none"
  )
  
  list(
    condition = condition,
    pair_df = pair_df,
    wilcox = wilcox_result,
    limma = limma_result
  )
}

proportion_results <- setNames(
  lapply(
    trained_conditions,
    run_proportion_paired_test
  ),
  trained_conditions
)

proportion_summary <- bind_rows(
  lapply(
    proportion_results,
    function(res) {
      
      if (is.null(res$wilcox)) {
        
        return(
          tibble(
            Condition = res$condition,
            N_subjects = nrow(res$pair_df),
            Mean_diff_pct_points = NA_real_,
            N_increased = NA_integer_,
            Wilcoxon_p = NA_real_,
            Limma_p = NA_real_
          )
        )
      }
      
      tibble(
        Condition = res$condition,
        N_subjects = nrow(res$pair_df),
        Mean_diff_pct_points = round(
          mean(
            res$pair_df$Difference_pct_points
          ),
          3
        ),
        N_increased = sum(
          res$pair_df$Difference_pct_points > 0
        ),
        Wilcoxon_p = res$wilcox$p.value,
        Limma_p = res$limma$P.Value[1]
      )
    }
  )
) %>%
  mutate(
    Panel_adj.P.Val = p.adjust(
      Limma_p,
      method = "BH"
    )
  )

kable(
  proportion_summary,
  digits = 4,
  caption = "Five-transcript CD8 percentage with panel-specific correction"
)
```
<!--
# 3. Pseudobulk expression of the requested marker genes

This analysis evaluates expression of the six prespecified genes in the pre-PMA CD8 population using subject-level pseudobulk RNA counts.
-->
```{r}
#| label: marker-expression-pseudobulk

cd8_pre <- subset(
  cd8,
  subset =
    Treatment == "Untreated" &
    LaneGroup %in% c(
      "untrained",
      trained_conditions
    )
)

DefaultAssay(cd8_pre) <- "RNA"

counts <- GetAssayData(
  cd8_pre,
  assay = "RNA",
  layer = "counts"
)

meta <- cd8_pre@meta.data

# Create one pseudobulk sample per subject x condition
meta$pseudobulk_id <- paste(
  meta$sCALL,
  meta$LaneGroup,
  sep = "_"
)

pb_ids <- unique(
  meta$pseudobulk_id
)

pb_counts <- sapply(
  pb_ids,
  function(id) {
    
    cells <- rownames(meta)[
      meta$pseudobulk_id == id
    ]
    
    Matrix::rowSums(
      counts[
        ,
        cells,
        drop = FALSE
      ]
    )
  }
)

# Make sure pseudobulk counts remain a matrix
pb_counts <- as.matrix(pb_counts)

colnames(pb_counts) <- pb_ids

# Pseudobulk metadata
pb_meta <- meta %>%
  select(
    pseudobulk_id,
    sCALL,
    LaneGroup
  ) %>%
  distinct()

pb_meta <- pb_meta[
  match(
    colnames(pb_counts),
    pb_meta$pseudobulk_id
  ),
  ,
  drop = FALSE
]

pb_meta$sCALL <- factor(
  pb_meta$sCALL
)

pb_meta$LaneGroup <- factor(
  pb_meta$LaneGroup,
  levels = c(
    "untrained",
    trained_conditions
  )
)

# Remove pseudobulk samples with missing subject or condition
valid_pb <- (
  !is.na(pb_meta$sCALL) &
  !is.na(pb_meta$LaneGroup)
)

pb_meta <- pb_meta[
  valid_pb,
  ,
  drop = FALSE
]

pb_counts <- pb_counts[
  ,
  valid_pb,
  drop = FALSE
]

cat(
  "Number of pseudobulk samples:",
  ncol(pb_counts),
  "\n"
)

#kable(
#  pb_meta,
#  caption = "Pseudobulk samples"
#)

# edgeR object
dge <- edgeR::DGEList(
  counts = pb_counts
)

# Filter genes with insufficient expression
keep <- edgeR::filterByExpr(
  dge,
  group = pb_meta$LaneGroup
)

dge <- dge[
  keep,
  ,
  keep.lib.sizes = FALSE
]

# TMM normalization
dge <- edgeR::calcNormFactors(
  dge
)

# Subject-paired design
design <- model.matrix(
  ~ sCALL + LaneGroup,
  data = pb_meta
)

cat(
  "Design matrix dimensions:",
  nrow(design),
  "x",
  ncol(design),
  "\n"
)

# voom + limma
v <- limma::voom(
  dge,
  design,
  plot = FALSE
)

fit <- limma::lmFit(
  v,
  design
)

fit <- limma::eBayes(
  fit
)
```

<!--
# 4. Six prespecified marker genes
# 4. Six prespecified biologically targeted marker genes

The six genes evaluated here (**TCF7, PDCD1, SLAMF6, CD28, CCR7, and SELL**) were selected **a priori** based on their established biological relevance to the stem-like CD8 T-cell phenotype of interest. This is therefore a targeted hypothesis-driven analysis rather than an exploratory transcriptome-wide screen.

For this targeted analysis, **nominal p-values are used to evaluate the prespecified gene-level hypotheses and are not subjected to FDR correction for the primary interpretation**. Effect sizes (log-fold changes) and consistency across subjects should also be considered when interpreting the results.
-->

```{r}
#| label: targeted-marker-results

extract_targeted_markers <- function(
  coef_name,
  condition_label
) {
  
  all_results <- limma::topTable(
    fit,
    coef = coef_name,
    number = Inf,
    sort.by = "none"
  )
  
  target <- all_results[
    rownames(all_results) %in% genes_to_test,
    ,
    drop = FALSE
  ]
  
  target <- target[
    match(
      genes_to_test,
      rownames(target)
    ),
    ,
    drop = FALSE
  ]
  
  tibble(
    Gene = genes_to_test,
    Comparison = condition_label,
    logFC = target$logFC,
    AveExpr = target$AveExpr,
    P.Value = target$P.Value
  )
}

marker_results <- bind_rows(
  
  extract_targeted_markers(
    "LaneGroupSYKi",
    "SYKi vs untrained"
  ),
  
  extract_targeted_markers(
    "LaneGroupMDP",
    "MDP vs untrained"
  ),
  
  extract_targeted_markers(
    "LaneGroupBG",
    "BG vs untrained"
  )
  
)

marker_results_display <- marker_results %>%
  mutate(
    logFC = round(logFC, 3),
    AveExpr = round(AveExpr, 3),
    P.Value = signif(P.Value, 4)
  )

kable(
  marker_results_display,
  digits = 4,
  caption = "Targeted analysis of six prespecified stem-like CD8 marker genes. Nominal p-values are reported."
)
```

# 5. Marker-expression plot

Red bars indicate **nominal p < 0.05** for the prespecified targeted genes. Because these genes were selected a priori based on the biological hypothesis, FDR correction is not used as the significance criterion in this targeted analysis.

```{r}
#| label: marker-expression-plot
#| fig-width: 9
#| fig-height: 5

marker_plot_data <- marker_results %>%
  mutate(
    Comparison = factor(
      Comparison,
      levels = c(
        "SYKi vs untrained",
        "MDP vs untrained",
        "BG vs untrained"
      )
    ),
    
    Significant = (
      !is.na(P.Value) &
      P.Value < 0.05
    )
  )

ggplot(
  marker_plot_data,
  aes(
    x = Gene,
    y = logFC,
    fill = Significant
  )
) +
  geom_col() +
  
  geom_hline(
    yintercept = 0,
    linewidth = 0.3
  ) +
  
  facet_wrap(
    ~ Comparison
  ) +
  
  scale_fill_manual(
    values = c(
      `TRUE` = "#c0392b",
      `FALSE` = "grey70"
    ),
    guide = "none"
  ) +
  
  labs(
    title = "Pseudobulk expression of prespecified stem-like CD8 markers",
    subtitle = "Red indicates nominal P < 0.05; genes were selected a priori based on known biology",
    x = NULL,
    y = "logFC (trained vs. untrained)"
  ) +
  
  theme_bw(
    base_size = 12
  ) +
  
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```
# 6. Individual-subject expression of the six targeted genes

The pseudobulk differential-expression analysis estimates the overall training-associated effect while accounting for subject. The plots below show the underlying expression values for each individual subject. Each line connects samples from the same subject across the four pre-PMA conditions.

This visualization is important for determining whether the direction of a gene-expression difference is consistent across subjects or is primarily driven by an individual subject.

```{r}
#| label: individual-subject-gene-expression
#| fig-width: 12
#| fig-height: 8


# Extract normalized pseudobulk expression from the voom object
voom_expression <- as.data.frame(v$E)

# Keep the six prespecified genes
individual_expression <- voom_expression[
  rownames(voom_expression) %in% genes_to_test,
  ,
  drop = FALSE
]

# Add gene names
individual_expression$Gene <- rownames(individual_expression)

# Convert to long format
individual_expression_long <- individual_expression %>%
  pivot_longer(
    cols = -Gene,
    names_to = "pseudobulk_id",
    values_to = "logCPM"
  ) %>%
  left_join(
    pb_meta %>%
      mutate(
        pseudobulk_id = as.character(pseudobulk_id)
      ),
    by = "pseudobulk_id"
  ) %>%
  mutate(
    LaneGroup = factor(
      LaneGroup,
      levels = c(
        "untrained",
        "SYKi",
        "MDP",
        "BG"
      )
    ),
    Gene = factor(
      Gene,
      levels = genes_to_test
    )
  )

# Show the data used for the plot
#kable(
#  individual_expression_long %>%
#    arrange(Gene, sCALL, LaneGroup),
#  digits = 3,
#  caption = "Individual-subject pseudobulk expression (voom logCPM)"
#)

# Individual-subject plot
ggplot(
  individual_expression_long,
  aes(
    x = LaneGroup,
    y = logCPM,
    group = sCALL
  )
) +
  geom_line(
    alpha = 0.65,
    linewidth = 0.7
  ) +
  geom_point(
    aes(shape = sCALL),
    size = 3
  ) +
  facet_wrap(
    ~ Gene,
    scales = "free_y",
    ncol = 3
  ) +
  labs(
    title = "Individual-subject expression of targeted CD8 markers",
    subtitle = "Pre-PMA pseudobulk expression; lines connect the same subject",
    x = NULL,
    y = "Expression (voom logCPM)",
    shape = "Subject"
  ) +
  theme_bw(
    base_size = 12
  ) +
  theme(
    strip.text = element_text(
      face = "bold"
    ),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    legend.position = "bottom"
  )
```



# 7. Summary

At the subject level, SYKi produced the most consistent increase in the RNA-defined stem-like/memory CD8 population, with an increase observed in 4 of 5 subjects and an average increase of 1.31 percentage points compared with untrained. Although this did not reach statistical significance with only five subjects (limma p=0.075; Wilcoxon p=0.106), the consistency and direction of the SYKi response support further investigation. 
Targeted pseudobulk analysis of the six stem-like CD8-associated genes was also performed to determine whether training is associated with changes in their expression. The six genes evaluated in this analysis (**TCF7, PDCD1, SLAMF6, CD28, CCR7, and SELL**) were prespecified based on their known relevance to the stem-like CD8 T-cell phenotype rather than selected from an exploratory genome-wide analysis.

Accordingly, the targeted pseudobulk analysis reports **nominal p-values** for the individual gene-level hypotheses. A nominal **P < 0.05** is used to highlight results in the figure. The magnitude and direction of the log-fold change and consistency of the response across subjects should be considered together with the p-values.

The five-transcript cell population is separately defined by simultaneous RNA detection of **TCF7, SLAMF6, CD28, CCR7, and SELL** within the same Monaco-defined CD8 cell. Subject-level proportions are used so that individual cells are not treated as independent biological replicates.
